The brightness and spatial distributions of terrestrial radio sources

### 2013-04-09 articles/Article-RFI-distribution/article-rfi-distribution.tex
Article fixes

@@

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after long integrations of radio observations if they are consistently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe. We explore the RFI situation at 30--163~MHz by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology. The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law implies that the current LOFAR EoR observations should be severely RFI limited if the strength of RFI sources remain strong after integration, i.e., they are observed as coherent sources. This is in contrast with actual observations, which almost reach the thermal noise and are thought not to be limited by RFI. Therefore, we conclude that it is unlikely that there are undetected RFI sources that will become visible after long-time integration. Consequently, it is likely that RFI will not prevent an EoR detection with LOFAR.
\end{abstract}

@@

\section{Introduction}
Radio astronomy concerns itself with the observation of radiation from celestial sources at radio wavelengths. However, astronomical radio observations can be affected by radio-frequency interference (RFI), which might make it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}. The careful management of spectrum allocation and the construction of radio-quiet zones help to limit the number of harmful transmitters. If harmful RFI is observed nevertheless, the use of RFI mitigation methods can sometimes clean the data sufficiently to allow succesful calibration and imaging. Many techniques have been designed to mitigate the effects of RFI, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, post-correlation-rfi-classification}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.

Typical radio observations record a few hours of data, and the results are integrated. In these cases, excising only the interference that is apparent and thus above the noise often suffices, i.e., the observation can still reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, weak interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be persistently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. Ground-based Cosmic Microwave Background (CMB) experiments are another class of experiments involving long integration times (e.g., \citealt{ska-cmb-subrahmanyan}). For these experiments, it is important to know the possible effect of low-level interference on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, van Haarlem et al. 2013, \textit{submitted}).
% TODO correct reference if van Haarlem et al is published
\nocite{lofar-overview-2013}

This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods. Our approach is similar to the radio-source-count analyses that are used in cosmology \citep{condon-cosm-evol-of-radio-sources}, also named $\log N$ -- $\log S$ analyses, where $N$ and $S$ refer to the celestial source count and brightness respectively. The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe. We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of estimating their full spatial and brightness distributions. This results in a better insight into the effects of low-level interference and allows one to simulate the effects of interference more accurately.
% A distribution can be expressed either as the probability density function (PDF) or as cumulative density function (CDF). This work uses PDFs, because they are better suited for visualizing the parameters we are interested in, although the two distributions basically convey the same information.

@@

This paper is organised as follows: in Sect.~\ref{sec:distribution-prediction}, we calculate a model for terrestrial interfering source distribution based on various assumptions. Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse brightness histograms of LOFAR data. This is followed in Sect.~\ref{sec:dist-data} by a short description of the two LOFAR data sets that have been used to perform the experiment. The results of analysing the sets are presented in Sect.~\ref{sec:dist-results}. Finally, in Sect.~\ref{sec:dist-discussion} the results are discussed and conclusions are drawn.

@@

In time-frequency space, interfering sources can have complex structures. They can also be intermittent and different sources might overlap in time-frequency space. An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a dynamic spectrum. Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in time-frequency space. It is often not clear what constitutes a single interfering source, hence it is hard to count individual sources. Instead, we will count the number of times a given brightness occurs in time-frequency space. This --- as well as many other effects --- will of course influence the distribution. If sources overlap in the time-frequency space, the situation is somewhat similar to the case where multiple unresolved celestial radio sources in the reception pattern of a telescope only allow observation of a sum of sources. However, in that case it is still possible to validate radio source models by comparing $\log N$ -- $\log S$ histograms \citep{scheuer-analysing-faint-stars}.

@@

\begin{figure*}
\begin{center}
\includegraphics[height=8cm]{img/TF-L43786_L220-rfi-example-xaxistext.pdf}
\caption{A small part of an observation displayed in a dynamic spectrum. The features with significantly higher values are caused by interference. Some of these have a constant frequency, while others are more erratic.}
\label{fig:dist-rfi-example}
\end{center}
\end{figure*}

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the term ``rate density''. We will now start by estimating the rate density function of ground-based interfering sources. Consider an interfering point source of strength $I$ that denotes the transmitting power normalized by the observational channel resolution (e.g., measured in W/Hz). This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects. The antennas are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{4\pi r_1 r_2},
\end{equation}
with (real-valued amplitude) gains $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. The formula represents a spherically propagating wave in free space. We will limit our analysis to cross-correlated antennas; the auto-correlations will be ignored.

We assume that the source observed is fully coherent, but a possible de-coherence factor can be absorbed in the gains. Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition. With a frequency resolution $\Delta\nu=$ 0.76~kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c (2 \pi \Delta \nu)^{-1} \approx$ 50~km. Because the velocity resolution of LOFAR is 1.5~km/s at 150~MHz, and larger at lower frequencies, a Doppler frequency shift due to movement of the source will only be significant if its velocity is at least 1.5~km/s relative to the antennas. Since the relative velocities to different antennas in the array will be similar for such high-velocity transmitters (i.e., satellites), there will be hardly any decorrelation because of Doppler shifting.

Although two antennas do not necessarily observe the same RFI sources, for source count analysis we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution. Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and interferometric gain $g = g_1 g_2$ as
\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
 S(r) = \frac{Ig}{4\pi r^2}.
\end{equation}

Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in the local two-dimensional horizontal plane. This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights. Beyond some distance, the Earth will partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid. Using these assumptions, we can express the expected cumulative rate density of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=\rho \pi r^2,
\end{equation}
for some constant $\rho$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.

We need $F_{\textrm{distance} \ge r}(r)$, the complement of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance. Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources. Because $r$ is restricted, $N$ is finite.

The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can now be calculated with
\begin{align} \label{eq:cumulative-rfi-distribution}
 F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{R}(S))\\
\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{4\pi S})\\
\notag = & N-\frac{\rho Ig}{4S}
\end{align}
where $\mathcal{R}(S)=\mathcal{S}^{-1}$, the inverse of $\mathcal{S}$, i.e., the function that returns the distance $r$ for a given amplitude $S$. Finally, the rate density can be calculated by taking the derivative,
\begin{align} \label{eq:two-dimensional-distribution}
 f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{\rho Ig}{4S^2}.
\end{align}

@@

A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. The original model is based on urban areas, but corrections are given by Hata for suburban areas with a lower population density and for open areas. Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennas in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \frac{1}{10} 10^{L_p} = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation} \label{eq:zeta-definition}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{Ig}{4\pi \zeta r^\eta},
\end{equation}
with $\eta = 4.49 - 0.655 \log_{10} h_b$ as seen in Eq.~\eqref{eq:propagation-loss-factor}, one finds the rate density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{4\pi\zeta S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{4\pi\zeta S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram. According to Hata's model, this is valid for the range 30--200 m. The trend of the slope will not continue indefinitely when increasing the height further. Instead it will converge to a $-2$ slope, which corresponds to free-space propagation.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

\subsection{Thermal noise contribution} \label{sec:histogram-noise}
The full measured distribution will consist of the power-law distribution combined with that of the thermal noise and the celestial signal. For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources. For example, the strongest apparent source in the NCP EoR field is around 5 Jy (Yattawata et al. 2013, \textit{in press}). % TODO fill in details.
The standard deviation of the noise, however, is around 100~Jy on highest LOFAR resolutions, and will have a larger contribution on the histogram.

@@

\begin{figure}
\begin{center}\hspace{-5mm}\includegraphics[width=8cm]{img/plot-rayleigh-and-rfi-trimmed}
\caption{The Rayleigh and power-law distributions in a log-log plot. The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2 over the range it is defined. The slope of the Rayleigh distribution in the limit of the origin is 1. Its maximum occurs where the amplitude value equals its mode $\sigma$, which is 1 in this example. For higher amplitudes, its slope decreases exponentially.}
\label{fig:rayleigh-and-rfi-distributions}
\end{center}
\end{figure}

@@

For completeness we show the formula for sampling apparent source strengths that include propagation effects according to Hata's model, which can be used for sampling realistic source amplitudes in simulations. This can be derived by integration, scaling and inversion of the rate density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
\begin{equation} \label{eq:sample-with-propagation}
\tilde{S}_{\textrm{RFI}} \leftarrow \frac{I g}{\zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
\end{equation}
Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 < x_u,y_u \le 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.

\subsection{Parameter variability} \label{sec:parameter-variability}
In reality, the parameters $\rho$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant, but can change over time and frequency. Therefore, they are stochastic variables. However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.

>>>>

\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}

@@

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after \editmark{long integrations} of radio observations if they \editmark{are consistently} present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe. We explore the RFI situation \editmark{at 30--163~MHz} by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology. The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law \editmark{implies} that the current LOFAR EoR observations should be severely RFI limited if the \editmark{strength of RFI sources remain strong after integration}. This is in contrast with actual observations, which almost reach the thermal noise and are thought not to be limited by RFI. Therefore, we conclude that it is \editmark{unlikely that there are undetected RFI sources that will become visible after long integration. Consequently, it is likely} that RFI will not prevent an EoR detection with LOFAR.
\end{abstract}

@@

\section{Introduction}
Radio astronomy concerns itself with the observation of radiation from celestial sources at radio wavelengths. However, astronomical radio observations can be affected by radio-frequency interference (RFI), which might make it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}. \editmark{The careful management of spectrum allocation and the construction of radio-quiet zones help to limit the number of harmful transmitters. If harmful RFI is observed nevertheless, the use of RFI mitigation methods can sometimes clean the data sufficiently to allow succesful calibration and imaging.} Many techniques have been designed to mitigate the effects of RFI, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, post-correlation-rfi-classification}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.

Typical radio observations record a few hours of data, and the results are integrated. In these cases, excising only the interference that is apparent and thus above the noise often suffices, i.e., the observation can still reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, weak interference caused by stationary \editmark{RFI sources} might not manifest itself above the noise in individual observations, but might be \editmark{persistently} present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. Ground-based Cosmic Microwave Background (CMB) experiments are another class of experiments involving long integration times (e.g., \citealt{ska-cmb-subrahmanyan}). For these experiments, it is important to know the possible effect of low-level interference on the data, as these might overshadow or alter the signal of interest.

In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, \citealt{lofar-overview-2013}, \textit{submitted}). \editmark{The LOFAR EoR project aims to detect the redshifted 21-cm signals from the EoR using the LOFAR HBA antennas (115--190~MHz, $z_\textrm{HI}$=11.4--6.5). Several fields will be observed over 100~nights, to achieve sufficient sensitivity to allow the signal extraction. An EoR calibration pipeline has been designed that solves for ionospheric and instrumental effects in approximately hundred directions using the SAGE algorithm \citep{sage-calibration-ii}. Initial results from commissioning observations show that in a single night the thermal noise level can almost be reached \citep{ncp-eor-yatawatta}.}
% TODO correct reference if van Haarlem et al is published

This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods. Our approach is similar to the radio-source-count analyses that are used in cosmology \citep{condon-cosm-evol-of-radio-sources}, also named $\log N$ -- $\log S$ analyses, where $N$ and $S$ refer to the celestial source count and brightness respectively. The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe. We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of estimating their full spatial and brightness distributions. This results in a better insight into the effects of low-level interference and allows one to simulate the effects of interference more accurately.

@@

This paper is organised as follows: in Sect.~\ref{sec:distribution-prediction}, we calculate a model for terrestrial interfering source distribution based on various assumptions. Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse \editmark{brightness histograms of LOFAR data}. This is followed in Sect.~\ref{sec:dist-data} by a short description of the two LOFAR data sets that have been used to perform the experiment. The results of analysing the sets are presented in Sect.~\ref{sec:dist-results}. Finally, in Sect.~\ref{sec:dist-discussion} the results are discussed and conclusions are drawn.

@@

In \editmark{time-frequency space}, interfering sources can have complex structures. They can also be intermittent and different sources might overlap in \editmark{time-frequency space}. An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a \editmark{dynamic spectrum}. Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in \editmark{time-frequency space}. It is often not clear what constitutes a single interfering source, hence it is hard to count individual sources. Instead, we will count the number of times a given brightness occurs in time-frequency space. This --- as well as many other effects --- will of course influence the distribution. If sources overlap in the time-frequency space, the situation is somewhat similar to the case where multiple unresolved celestial radio sources in the reception pattern of a telescope only allow observation of a sum of sources. However, in that case it is still possible to validate radio source models by comparing $\log N$ -- $\log S$ histograms \citep{scheuer-analysing-faint-stars}.

@@

\begin{figure*}
\begin{center}
\includegraphics[height=8cm]{img/TF-L43786_L220-rfi-example-xaxistext.pdf}
\caption{A \editmark{dynamic spectrum} of a small part of an observation. The features with significantly higher values are caused by interference. Some of these have a constant frequency, while others are more erratic.}
\label{fig:dist-rfi-example}
\end{center}
\end{figure*}

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the term \editmark{``rate density''}. We will now start by estimating the rate density function of ground-based interfering sources. Consider an interfering point source of strength $I$ that denotes \editmark{the transmitting power normalized by the observational channel resolution (e.g., measured in W/Hz)}. This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects. The antennas are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\editmark{\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{4\pi r_1 r_2},
\end{equation}}
with (real-valued amplitude) gains $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. The formula represents a spherically propagating wave in free space. We will limit our analysis to cross-correlated antennas; the auto-correlations will be ignored.

We assume that the source observed is fully coherent, but a possible de-coherence factor can be absorbed in the gains. Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition. \editmark{With a frequency resolution $\Delta\nu=$ 0.76~kHz}, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c (2 \pi \Delta \nu)^{-1} \approx$ 50~\editmark{km}. \editmark{Because the velocity resolution of LOFAR is 1.5~km/s at 150~MHz, and larger at lower frequencies, a Doppler frequency shift due to movement of the source will only be significant if its velocity is at least 1.5~km/s relative to the antennas. Since the relative velocities towards different antennas in the array will be similar for such high-velocity transmitters (i.e., satellites), there will be hardly any decorrelation because of Doppler shifting.}

Although two antennas do not necessarily observe the same RFI sources, for source count analysis we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution. Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and interferometric gain $g = g_1 g_2$ as
\editmark{\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
 S(r) = \frac{Ig}{4\pi r^2}.
\end{equation}}

Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in the local two-dimensional horizontal plane. This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights. Beyond some distance, the Earth will partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid. Using these assumptions, we can express the expected cumulative \editmark{rate density} of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=\rho \pi r^2,
\end{equation}
for some constant $\rho$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.

We need \editmark{$F_{\textrm{distance} \ge r}(r)$, the complement} of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance. Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources. Because $r$ is restricted, $N$ is finite.

The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can now be calculated with
\editmark{\begin{align} \label{eq:cumulative-rfi-distribution}
 F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{R}(S))\\
\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{4\pi S})\\
\notag = & N-\frac{\rho Ig}{4S}
\end{align}}
where $\mathcal{R}(S)=\mathcal{S}^{-1}$, the inverse of $\mathcal{S}$, i.e., the function that returns the distance $r$ for a given amplitude $S$. Finally, the \editmark{rate density} can be calculated by taking the derivative,
\editmark{\begin{align} \label{eq:two-dimensional-distribution}
 f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{\rho Ig}{4S^2}.
\end{align}}

@@

A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. The original model is based on urban areas, but corrections are given by Hata for suburban areas with a lower population density and for open areas. Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss \editmark{between two ground-based antennas}:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennas in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\editmark{\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \frac{1}{10} 10^{L_p} = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}}
with
\begin{equation} \label{eq:zeta-definition}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\editmark{\begin{equation}
 S(r) = \frac{Ig}{4\pi \zeta r^\eta},
\end{equation}}
with $\eta = 4.49 - 0.655 \log_{10} h_b$ as seen in Eq.~\eqref{eq:propagation-loss-factor}, one finds the rate density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
\editmark{\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{4\pi\zeta S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{4\pi\zeta S} \right)^{2/\eta}.
\end{align}}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram. According to Hata's model, this is valid for the range 30--200 m. The trend of the slope will not continue indefinitely when increasing the height further. Instead it will converge to a $-2$ slope, which corresponds to free-space propagation.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

\subsection{Thermal noise contribution} \label{sec:histogram-noise}
The full measured distribution will consist of the power-law distribution combined with that of the thermal noise and the celestial signal. For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong \editmark{celestial} sources. For example, the strongest apparent \editmark{celestial} source in the NCP EoR field is around 5 Jy \citep{ncp-eor-yatawatta}.
The standard deviation of the noise, however, is around 100~Jy on highest LOFAR resolutions, and will have a larger contribution on the histogram.

@@

\begin{figure}
\begin{center}\hspace{-5mm}\includegraphics[width=8cm]{img/plot-rayleigh-and-rfi-trimmed}
\caption{The Rayleigh and power-law distributions in a log-log plot. The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2 \editmark{over the range it is defined}. The slope of the Rayleigh distribution in the limit of the origin is 1. Its maximum occurs where the amplitude value equals its mode $\sigma$, which is 1 in this example. For higher amplitudes, its slope decreases exponentially.}
\label{fig:rayleigh-and-rfi-distributions}
\end{center}
\end{figure}

@@

For completeness we show the formula for sampling apparent source strengths that include propagation effects according to Hata's model, which can be used for sampling realistic source amplitudes in simulations. This can be derived by integration, scaling and inversion of the \editmark{rate density} function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
\begin{equation} \label{eq:sample-with-propagation}
\tilde{S}_{\textrm{RFI}} \leftarrow \frac{I g}{\zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
\end{equation}
Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 < x_u,y_u \le 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.
